
\section{Memoization system}
There is a strong trend nowadays in scientific computations~-- the size of data to be processed increases faster than available computing resources. Different optimization techniques are used to overcome this problem. For example, a memoization technique, which is as follows~-- a result~$R$ that has been computed once by a function~$F$ with the argument~$A$ is stored in memory (together with $A$). Then on consequent calls of $F$ with~$A$, $R$ is retrieved from memory, hence the actual call to~$F$ is omitted. This optimization turns out to be particularly useful when a result of~$F$ takes a long time to be computed (relatively to a time needed to retrieve $R$ from memory), and $F$ is usually called only with a small subset of arguments. Sometimes this technique is also referred to as a caching.

A target audience for this kind of memoization systems is scientific application developers.

The concept of a numerical database, introduced by Park, Draayer, and Zheng\cite{park90}, describes one of such memoization systems. Its main purpose is to ``store and retrieve valuable intermediate information so costly redundant calculations can be avoided.'' Authors also presented a possible implementation, that is based on a weighted search tree.

The main goal of this thesis is to explore different look-up data structures, especially those supporting a concurrent access, which can be used in place of the weighted search tree and compare their performance. Then, the most promising data structures will be packed in a programming library, written in C++ programming language.


\section{Thesis structure}
The remainder of this thesis is structured as follows:
 \begin{description}
 \item[In \cref{ch:pre}] some basic programming concepts and data structures are described. The rest of the thesis relies on these terms.
 \item[In \cref{ch:numdb}] the contributions of~\cite{park90} are recapitulated~-- the numerical database concept and the weighted search tree are explained in~detail.
 \item[In \cref{ch:alt}] all of the main contributions of this thesis are presented. \\ Specifically, the overview of different eviction policies, other than what have been proposed in~\cite{park90}, is provided. Furthermore, alternative data structures, which can be used as a base for the numerical database, are discussed.

 \item [\Cref{ch:impl}] contains general information as well as a more in-depth overview of the \numdbname{} program library. This library provides an implementation of several containers (including concurrent containers), discussed in chapters 2 and 3.

 \item [In \cref{ch:bench}] the results of the performance comparison of the containers are presented and concluded.

 \item [In conclusion] the outcomes of this thesis are recapitulated.

\end{description}
